---
title: "Lab 8 part 2"
author: "Max Settineri"
date: "2023-03-02"
output: html_document
---

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning= FALSE, message = FALSE)
library(tidyverse)
library(here)
library(lubridate)
library(tsibble)
library(feasts)
library(slider)
```

## Read in data

```{r}
toolik <- read_csv(here('data/toolikweather.csv')) 

toolik_ts <- toolik %>% 
  mutate(date = mdy(date)) %>% 
  as_tsibble(key = NULL, index = date)

ggplot(data = toolik_ts, aes(x = date, y = mean_airtemp)) +
  geom_line() +
  labs(x = "Date",
       y = "Mean daily air temperature (Celsius)\n at Toolik Station")
```

## Use `index_by()` to aggregate time series by increments

```{r}
toolik_month <- toolik_ts %>% 
  index_by(yr_mo = ~yearmonth(.)) %>% 
  summarize(monthly_mean_temp = mean(mean_airtemp, na.rm = TRUE))

ggplot(toolik_month, aes(x = yr_mo, y = monthly_mean_temp)) +
  geom_line()

toolik_month %>% 
  ggplot(aes(x = year(yr_mo), y = monthly_mean_temp)) +
  geom_line() +
  facet_wrap(~ month(yr_mo, label = TRUE)) +
  labs(x = 'Year', y = 'Annual mean air temp (C)',
       title = 'Toolik Station mean annual air temperature',
       subtitle = '1988-2018',
       caption = 'Source: Shaver, G. 2019. A multi-year daily weather file
       for the Toolik Field Station at Toolik Lake, AK, starting 1988 to present. 
       Ver. 4. Environmental Data Inititative')
```

### Group by other increments in 'index_by'

```{r}
toolik_annual <- toolik_ts %>% 
  index_by(yearly = ~year(.)) %>% 
  summarize(annual_airtemp = mean(mean_airtemp, na.rm = TRUE))

ggplot(toolik_annual, aes(x = yearly, y = annual_airtemp)) +
  geom_line() 

toolik_weekly <- toolik_ts %>% 
  index_by(weekly = ~yearweek(.)) %>% 
  summarize(weekly_airtemp = mean(mean_airtemp, na.rm = TRUE))

ggplot(toolik_weekly, aes(x = weekly, y = weekly_airtemp)) +
  geom_line()
```

## Use filter_index to grab different date/time combos

```{r}
toolik_ts %>% 
  filter_index('2000-06' ~ '2001-10')

toolik_ts %>% 
  filter_index('2006-04-10' ~ '2006-05-15')

toolik_ts %>% 
  filter_index('2017-12-20' ~ .) # from date till end of data set
```

## Exploring seasonality

```{r}
toolik_ts %>% 
  filter_index('2014' ~ .) %>% 
  feasts::gg_season(y = mean_airtemp)

toolik_month %>% 
  gg_season(y = monthly_mean_temp) +
  labs(x = 'year', y = 'mean monthly air temp (C)')
```

## Seasonal subseries plots

```{r}
toolik_month %>% 
  gg_subseries(y = monthly_mean_temp)
```

## Moving averages in tsibbles

We'll use the `slider` package to find moving (or rolling) averages for different window sizes. 

The general structure will tend to be something like: 

`df %>% slide(variable, function, .before = , .after = )`

Let's make a test vector just so we can see how this works: 

```{r}
set.seed(2023)
test<- rnorm(100, mean = 40, sd = 10)

### Show the series based on values +2 and -2 from each observation
### Use ~.x to show the windows
w05 <- slide(test, ~.x, .before = 2, .after = 2)

### Change that to a function name to actually calculate something for each window
### Note that I add `as.numeric` here, since the outcome is otherwise a list
w05 <- as.numeric(slide(test, mean, .before = 2, .after = 2))

### Find the mean value of a window with n = 11, centered:
w11 <- as.numeric(slide(test, mean, .before = 5, .after = 5))

### Find the mean value of a window with n = 19, centered:
w19 <- as.numeric(slide(test, mean, .before = 9, .after = 9))

### Plot these together: 
combo <- data_frame(time = seq(1:100), test, w05, w11, w19) %>%
  pivot_longer(cols = test:w19, names_to = 'series', values_to = 'value')


ggplot(data = combo) +
  geom_line(aes(x = time, y = value, color = series)) +
  scale_color_manual(values = c('grey70', 'red', 'orange', 'purple')) +
  theme_minimal()
```

## Find moving average on Toolik time series

```{r}
roll_toolik_15 <- toolik_ts %>% 
  mutate(ma_15d = as.numeric(slide(toolik_ts$mean_airtemp, mean, 
                                   .before = 7, .after = 7)))

roll_toolik_61 <- toolik_ts %>% 
  mutate(ma_61d = as.numeric(slide(toolik_ts$mean_airtemp, mean, 
                                   .before = 30, .after = 30)))


ggplot() +
  geom_line(data = toolik_ts, aes(x = date, y = mean_airtemp), 
            size = 0.2, color = "gray") +
  geom_line(data = roll_toolik_15, aes(x = date, y = ma_15d), 
            color = "orange") +
  geom_line(data = roll_toolik_61, aes(x = date, y = ma_61d), 
            color = "blue") +
  theme_minimal()

```

## Autocorrelation function

We'll look at outcomes for both daily lags (yikes) and monthly lags (cool).

```{r}
toolik_ts %>%
  ACF(mean_airtemp) %>%
  autoplot()

toolik_month %>% 
  ACF(monthly_mean_temp) %>% 
  autoplot()
```

## Decomposition

Here we will use STL decomposition (Seasonal, Trend, and Loess) decomposition. You can read about the advantages of STL decomposition here: https://otexts.com/fpp2/stl.html.


```{r}
toolik_dec <- toolik_month %>% 
  model(STL(monthly_mean_temp ~ season(window = Inf)))

components(toolik_dec) %>% autoplot()
```